---
title: "How many patients do you need for your study?"
license: "CC BY 4.0 for original teaching material; MIT for original code"
subtitle: "Sample size calculation tutorial"
author: "Katalin Tamási"
date: "5 October 2026"
lang: en
bibliography: references.bib
link-citations: true
nocite: '@paskevicius2017oep, @paskevicius2019open, @cronin2018oep, @orsini2024'
output:
  bookdown::gitbook:
    css: styles.css
    highlight: pygments
    code_folding: none
    split_bib: false
    config:
      sharing: false
      toc:
        collapse: section
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "##", message = FALSE,
  warning = TRUE, error = FALSE, fig.width = 7.5, fig.height = 4.8,
  fig.align = "center", out.width = "100%", dpi = 144)
options(width = 76, digits = 4)
source("../R/sample_size_functions.R")
source("../R/tutorial_helpers.R")
source("../R/teaching_cases.R")
source("../R/question_map.R")
```

# Welcome {-}

How many patients would make a study informative? This tutorial connects a clinical question to its assumptions, a sample size calculation, and the results that different repetitions of the study could produce.

The aims of this module are to understand the importance of recruiting an appropriate number of patients into your study and to be able to calculate the required sample size given reasonable assumptions when designing your study. This open edition retains the questions and clinical narrative of the original tutorial dated **14 May 2021** [@tamasi2021], with additional explanations and browser activities.

The [Introduction](#clinical-question) focuses on **two-arm randomized controlled clinical trials**. The treatment-comparison examples use two independent groups with equal allocation, a binary or continuous outcome, and a two-sided comparison. Later sections broaden the discussion to precision, prevalence surveys and pilot feasibility studies; [Where does my study fit?](#study-designs) guides you to methods for other designs. A single-mean example illustrates [Type I error rate](#alpha), [Type II error rate](#beta) and [power](#statistical-power) using a known population SD. All numerical cases are hypothetical teaching examples. Further background comes from lecture notes by @clayton2019 and the medical introductions by @kirkwood2003, @whitley2002, @pourhoseingholi2013 and @wang2020.

The presentation and three simulation activities are inspired by chapters 1, 11 and 15 of *Power Analysis with Superpower* [@superpowerbook2022]. The clinical explanations, activities and figures here are tailored to this tutorial.

Teaching material is shared under CC BY 4.0 and original code under MIT. You may reuse and adapt the material with attribution. The [teaching guide](../teaching/TEACHING_GUIDE.md), [case template](../case-studies/student_case_template.md) and [assessment rubric](../teaching/assessment_rubric.md) support classroom use. Public attribution or publication requires consent.

## How to use this tutorial {-}

Read the question, make a prediction, inspect the output, and explain what it means for the clinical study. Hover over a figure or table reference, or use its **Preview** button, to inspect it while reading. R code and raw console output are collapsed initially: select **R code and output** to inspect them. Formatted results, tables and figures remain visible. Each chapter offers HTML and PDF downloads; HTML retains the expandable explanations and code, while PDF expands the teaching explanations and keeps technical code hidden. Lines starting with `#` inside code are comments; `##` marks console output, and `[1]` identifies the first printed element of an R output. Select an amber **Common misconception** title to read its explanation.

Open [the tutorial and matching app together](http://127.0.0.1:8769/study/?activity=pain_one_many) for the guided activities. Each activity link loads the same case and assumptions used in the text. The [Sample size reasoning lab](http://127.0.0.1:8769/power_explorer/) is also available on its own. The prepared figures and results in this tutorial provide a route without the app. Other [clinical sample size tools](http://127.0.0.1:8769/apps.html) cover two proportions, two means and dropout.

For a reminder of a term, see [Key parameters and prerequisites](#glossary). If you are applying the material to your own project, begin with [Where does my study fit?](#study-designs) and [Plan your own study](#own-study). [Common mistakes and questions](#common-mistakes) collects explanations you can return to while working.

Need help finding your way around? See [Using the toolbar](#using-the-toolbar) below.

By the end, you should be able to:

- Choose a testing, estimation or fixed-resource planning goal and explain why it fits the question.
- Distinguish expected effects, planning effects and clinically meaningful thresholds.
- Justify assumptions and examine plausible alternatives, including when reusing an existing dataset.
- Explain Type I error rate, Type II error rate and power, and why one study can be inconclusive despite high planned power.
- Calculate approximate sample sizes for two means or two proportions, and plan the precision of a single-proportion estimate.
- Read power curves and distinguish per-group, total, analysable and recruitment counts.
- Interpret repeated-study rejection, coverage, precision and Monte Carlo uncertainty.
- Interpret an interval against zero and a clinical threshold, and recognise when a study design needs another planning method.
- Write a study justification that records evidence, assumptions, sensitivity, feasibility and limitations, and reproduce its calculations.


## Using the toolbar {#using-the-toolbar .unnumbered}

In the online book, the toolbar is the row of small icons at the top of the page. You can use it to find a chapter, search the tutorial, or make the text easier to read. The icons appear in this order:

| Icon | What it does | How to use it |
|:--|:--|:--|
| **Three horizontal lines** (contents) | Shows or hides the chapter list on the left. | Select a chapter or section to open it. On a phone, open this menu first to see the contents. Close it again with the same icon if it covers the text. |
| **Magnifying glass** (search) | Searches the text of the whole book. | Select it, type a word such as *dropout*, and select a result to open that passage. Press **Esc** to leave search. |
| **A** (reading settings) | Changes how the book looks on your screen. | Use the smaller or larger **A** to adjust text size. **Serif / Sans** changes the typeface; **White / Sepia / Night** changes the colour theme; **Spacing - / +** adjusts the line spacing. These choices do not change the content. |
| **Downward arrow** (download) | Opens the complete tutorial as one HTML page. | Select **Standalone HTML**. It opens in a new browser tab. To keep a local copy, use your browser's **Save page as** command. External links and some mathematical notation may still need internet access. |
| **i** (help) | Opens this guide to the toolbar. | Use it whenever you need a reminder. |

Table: (\#tab:toolbar-guide) Controls in the online tutorial toolbar.

To read in order, use the **left and right arrows at the sides of the page** to move to the previous or next chapter. The first page has no previous-chapter arrow, and the last page has no next-chapter arrow. On smaller screens, you may need to scroll down to find these arrows; the contents menu is another way to change chapters.

If you prefer the keyboard, press **Tab** to move between controls and **Enter** to activate one. When you are not typing in a text box, **S** shows or hides the contents, **F** opens search, and the **left / right arrow keys** move between chapters. Within search, use **up / down** to select a match and **Enter** to open it.

The downloaded, single-page HTML edition has its own table of contents rather than this book toolbar. Use your browser's **Find** command (**Ctrl+F** on Windows, **Command+F** on Mac) to search that edition.

# Introduction {#clinical-question}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Explain why sample size starts with the clinical question.
- Identify the assumptions needed before calculating.
- Distinguish expected effects, planning effects and clinical thresholds in a two-arm clinical trial.
:::

## The chronic-pain study: how many patients do we need?

Imagine planning a trial of a novel pain treatment against standard treatment. **Does the novel treatment increase the percentage of patients reporting pain relief at four weeks?**

We will randomly allocate patients to two independent groups of equal size. The primary outcome is pain relief at four weeks: yes or no. A small pilot found relief in **7 of 10** patients receiving the novel treatment and **3 of 10** receiving standard treatment: **70% versus 30%**, an observed difference of **40 percentage points**.

This looks promising. But twenty patients give us limited information. How many patients would make the main trial informative? We will build its plan step by step, then examine what different repetitions of that study could show.

This first section focuses on **two-arm randomized superiority trials**. Later chapters extend the reasoning to other outcomes, estimation, existing datasets and other study designs. All cases are hypothetical.

## What should this study establish?

A sample size calculation checks whether a planned study can answer its question. Too few patients can leave an important question unresolved; unnecessary recruitment also uses time and resources. For our pain trial, consider four connected questions:

- **Statistical:** can the study detect a treatment benefit and estimate its size precisely enough?
- **Clinical:** how much additional pain relief would matter to patients?
- **Ethical:** does the information the study could provide justify participants' burden and risk?
- **Practical:** can we recruit the required patients within the available time and budget?

The final plan should address all four. A statistical target alone does not establish clinical usefulness or feasibility.

## What assumptions do we need before calculating? {#how-do-you-determine-the-minimum-required-sample-size}

Start with the primary outcome, the expected results in each group, the planned analysis and the acceptable error rates. The map below gathers the questions to answer before choosing a sample size. Apply each question to the chronic-pain trial: we already know the outcome, the two independent groups and the equal allocation.

```{r study-questions-map, echo=FALSE, results='asis'}
knitr::asis_output(render_question_map())
```

Our examples use frequentist methods: two-sided significance tests for superiority and confidence intervals for estimation. Non-inferiority, equivalence and clustered trials require different hypotheses or calculations; [Where does my study fit?](#study-designs) helps identify these designs. For Bayesian planning, see [@kunzmann2021review].

## What do we expect, what will we plan for, and what matters clinically?

Suppose the pilot and other relevant evidence support an **expected benefit of 40 percentage points**. For the sample size calculation, use a **planning benefit of 30 percentage points**: 60% relief with the novel treatment versus 30% with standard treatment. Separately, consultation with patients and clinicians identifies a **clinical threshold of 20 percentage points**.

| Quantity | Value | Role in this study |
|:--|:--|:--|
| Expected benefit | 40 percentage points | Best current expectation, with uncertainty |
| Planning benefit | 30 percentage points: 60% versus 30% | Assumed true difference at which we evaluate power |
| Clinical threshold | 20 percentage points | Smallest benefit judged worthwhile by patients and clinicians |

Table: (\#tab:pain-planning-values) Three distinct decisions for the chronic-pain trial.

The three values in Table \@ref(tab:pain-planning-values) answer different questions. Using a smaller planning effect acknowledges uncertainty here; subtracting a fixed amount from every pilot estimate is not a general rule. Justify the choice from relevant evidence and examine plausible alternatives. Express absolute differences in **percentage points**, and avoid choosing an effect solely because a generic scale calls it small, medium or large [@caldwell2020].

<details class="misconception" id="misconception-pilot-effect">
<summary>Common misconception: the pilot reveals the true treatment effect</summary>

The pilot's 70% versus 30% relief rates are sample estimates and could change in another small sample. They inform our expectations, but do not establish the population effect. The expected benefit, effect used for planning and clinical threshold answer different questions and should each be justified.

</details>

## Does the pilot already answer the clinical question? {#critical-observed-difference}

The pilot's observed **40-percentage-point** difference exceeds the **20-percentage-point** clinical threshold. But the estimate is uncertain. Using a two-sided comparison of proportions gives a p-value of approximately **0.074**. With a Type I error rate of 5%, this does not provide sufficient evidence to reject the null hypothesis using the chosen test.

What would happen with one more response? Keep standard treatment at 3 of 10 patients, but increase novel-treatment relief from 7 to 8 of 10:

```{r pain-critical-p-values, warning=FALSE}
# Compare the pilot result with one additional treatment response
pain_p_value <- function(novel, standard, n) {
  prop.test(
    x = c(novel, standard),
    n = c(n, n),
    alternative = "two.sided",
    correct = FALSE
  )$p.value
}

round(c(
  "7 versus 3" = pain_p_value(7, 3, 10),
  "8 versus 3" = pain_p_value(8, 3, 10)
), 3)
```

```{r pain-critical-p-values-formatted, echo=FALSE}
knitr::kable(data.frame(`Novel treatment` = c("7/10 (70%)", "8/10 (80%)"), `Standard treatment` = "3/10 (30%)", `Observed difference` = c("40 percentage points", "50 percentage points"), `p-value` = formatC(c(pain_p_value(7,3,10), pain_p_value(8,3,10)), format="f", digits=3), check.names=FALSE),
  caption = "Pilot comparisons with seven or eight responses to the novel treatment.")
```

With 8 rather than 7 patients reporting relief, the result becomes statistically significant. For this particular comparison, the smallest observed difference giving $p<0.05$ is therefore $0.8-0.3=0.5$, or **50 percentage points**.

Notice that the clinical threshold remains 20 percentage points. What patients would consider a worthwhile improvement does not change because your pilot happens to contain only 10 patients per group.

<details class="misconception" id="misconception-significance">
<summary>Common misconception: statistical significance decides whether a treatment matters</summary>

A non-significant result does not establish that the treatments are equally effective. A significant result does not establish that the benefit matters to patients. Examine the estimated difference and its confidence interval against both zero and the clinical threshold.

</details>

*The calculation above uses an approximate test without continuity correction, matching the reasoning lab. With such small groups, an exact test may give a different result.*

### What happens if you recruit more patients?

Let's repeat this calculation for larger studies. At each sample size, keep the observed percentage reporting relief with standard treatment at 30%. Then find the smallest number reporting relief with the novel treatment that gives $p<0.05$.

```{r pain-critical-effect, echo=FALSE, warning=FALSE, fig.height=5.8, fig.cap="Smallest observed improvement giving p < 0.05 at each sample size, conditional on an observed control response of 30%. The two-sided score test uses no continuity correction. Points represent attainable patient counts; connecting lines guide the eye. The hypothetical clinical threshold is 20 percentage points. Counts are analysable patients before adjustment for dropout.", fig.alt="The smallest observed difference giving statistical significance falls from 50 percentage points at 10 patients per group to 14 at 100 per group. A horizontal line marks the clinical threshold of 20 percentage points. The pilot difference of 40 points at 10 patients per group lies below the significance threshold."}
pain_n <- seq(10, 100, by = 10)
pain_critical <- vapply(pain_n, function(n) {
  control <- as.integer(round(0.30 * n))
  treatment <- seq.int(control + 1L, n)
  p <- vapply(treatment, function(x) pain_p_value(x, control, n), numeric(1))
  min((treatment[p < 0.05] - control) / n)
}, numeric(1))

local({
  old_par <- par(mar = c(5, 5, 2, 1), las = 1)
  on.exit(par(old_par))
  plot(pain_n, 100 * pain_critical, type = "b", pch = 16, lwd = 2,
       col = "#17658A", ylim = c(0, 66), xaxt = "n",
       xlab = "Analysable patients per group (equal allocation)",
       ylab = "Observed difference in pain relief (percentage points)")
  axis(1, at = pain_n)
  abline(h = 20, col = "#A5500C", lty = 2, lwd = 2)
  points(10, 40, pch = 1, cex = 1.5, lwd = 2, col = "#303A43")
  legend("topright",
         c("Smallest observed difference with p < 0.05",
           "Clinical threshold: 20 percentage points",
           "Pilot: 7/10 versus 3/10"),
         col = c("#17658A", "#A5500C", "#303A43"),
         lty = c(1, 2, NA), pch = c(16, NA, 1),
         lwd = 2, bty = "n", cex = 0.8)
})
```

With 20 patients per group, an observed difference of 35 percentage points is sufficient for statistical significance under these conditions. With 50 patients per group, the corresponding difference is 20 percentage points. As the sample size increases, smaller observed differences can therefore become statistically significant.

Figure \@ref(fig:pain-critical-effect) shows how the observed significance boundary changes. The dashed line shows the clinically meaningful difference chosen for this example. With a larger sample, a result can be statistically significant even when the observed difference is smaller than this clinical threshold. You therefore need to consider the size of the difference and its confidence interval when interpreting the result.

### How does this relate to power?

So far, you have asked how large an **observed** difference must be to obtain $p<0.05$. When planning your main study, you need to ask a further question: *If the true treatment difference is 30 percentage points, how likely is your study to produce a statistically significant result?*

Different samples will give different observed differences, even when the true treatment effect remains the same. Some will be larger and others smaller. A power calculation takes this variation into account. We will examine this in the next chapter, after recording the assumptions for our study.

This illustration draws on the distinction between a critical observed effect and a sensitivity power analysis in @lakens2022, and section 1.7 of *Power Analysis with Superpower* [@superpowerbook2022].

::: {.activity}
Before continuing, consider a study that finds a statistically significant difference of 20 percentage points. **What would you still need to know before deciding whether the novel treatment offers a clinically meaningful benefit?**
:::

## Record the assumptions {#record-assumptions}

Write a study plan before choosing a sample size. Distinguish what you expect, what matters clinically, and the effect used to plan the study.

| Decision | What to make explicit |
|---|---|
| Question and outcome | Population, intervention, comparator, primary outcome and time point |
| Goal | Testing, estimation, fixed resources, or a stated combination |
| Effect scale | Treatment minus control; score units or percentage points; direction of benefit |
| Expected effect | Best current expectation, supporting evidence and uncertainty |
| Clinical threshold | Smallest benefit that matters to patients, with justification |
| Planning effect | Difference used in the power calculation, with justification |
| Variation or event rates | Common standard deviation for continuous outcomes, or control and treatment percentages |
| Sensitivity | Plausible alternative effects, standard deviations or event rates, with sources |
| Analysis and power | Planned test and confidence interval; sidedness, Type I error rate and target power |
| Precision | Desired full confidence interval width and confidence level |
| Allocation and counts | Allocation ratio; per-group and total analysable sample sizes |
| Recruitment | Expected losses, recruitment target, recruitment period and available resources |
| Intended conclusion | What the design could establish and what remains uncertain |

Table: (\#tab:planning-assumptions) Assumptions to record before planning the sample size.

The **planning effect** tells the calculation where to evaluate power. The **standard deviation** describes variation between individuals; the **standard error** describes sampling variation of an estimate. For binary outcomes, report each group's event rate as a percentage and their absolute difference in percentage points. [The glossary](#glossary) explains these terms.

::: {.activity}
**Record and justify.** [Open the assumptions report activity](http://127.0.0.1:8769/study/?activity=assumptions_report). Fill in the text boxes to build a study-plan report alongside your answers. Start with your clinical question and evidence; add the sample size and recruitment counts after using the calculation activities. You can load a hypothetical two-arm example, save your entries, and download the report. Check that the expected effect, planning effect and clinical threshold each have a clear role, and explain what your study could establish.
:::


Small pilot effects are uncertain. Following up only pilots whose estimated effect gives an affordable main study can introduce additional bias [@albers2018]. In this tutorial we examine plausible ranges explicitly. Safeguard power is an optional, more formal approach that uses uncertainty in the effect estimate when planning; it does not replace clinical judgement [@perugini2014].

::: {.activity}
**Predict and justify.** If the pilot had shown relief in 6 rather than 7 treated patients, should the clinical threshold change? Explain why a noisy pilot estimate and a judgement about meaningful benefit answer different questions.
:::

# What does statistical power tell you? {#power}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Explain Type I error rate, Type II error rate and power under a specified population scenario.
- Distinguish the outcome of one study from the performance of a design across repeated studies.
:::

Imagine repeating the same planned study many times, with the same sample size, analysis and specified true treatment effect. **Power is the percentage of those studies that would reject the null hypothesis given that the null hypothesis is false.** At 90% power, about 90 out of 100 repetitions would detect that effect if it exists, while about 10 would not. This describes the design's long-run performance. You do not need to conduct the trial repeatedly: calculations or simulations estimate this probability under the stated assumptions.

<details class="misconception" id="misconception-power-meaning">
<summary>Common misconception: 90% power means a 90% probability that the treatment works</summary>

Power assumes a particular true effect and describes how often the planned test would reject the null hypothesis across repeated studies. It does not tell us how likely that effect is to exist. A study with 90% power can still produce an inconclusive result.

</details>

We now return to the planning question: **if pain relief is 60% with the novel treatment and 30% with standard treatment, how often would studies of a given size detect the difference?** Power describes this repeated-study performance under the specified assumptions.

## Hypotheses and error probabilities

Let $\delta$ denote the signed population difference, treatment minus control. For an ordinary two-sided superiority test:

\begin{equation}
H_0: \delta=0, \qquad H_1: \delta\ne0.
(\#eq:hypotheses)
\end{equation}

In the chronic-pain case, planning for a 30-percentage-point difference ($\delta=0.30$) evaluates power at that particular effect. It does not change the alternative to $|\delta|\geq0.30$. The expected benefit remains 40 percentage points and the clinical threshold remains 20 percentage points. Rejecting zero alone does not show that this clinical threshold has been exceeded.

```{r error-table, echo=FALSE}
knitr::kable(data.frame(
  Decision = c("Reject the null", "Do not reject the null"),
  `Null is true` = c("Type I error", "Correct non-rejection"),
  `Specified alternative is true` = c("Power (1 - Type II error rate)", "Type II error"),
  check.names = FALSE), caption = "Testing decisions across repeated studies.")
```

**Type I error** means rejecting a true null hypothesis. **Type II error** means failing to reject a false null hypothesis. **Power** is the probability of rejection under a specified alternative, design and analysis [@whitley2002; @superpowerbook2022].

<details class="misconception" id="misconception-Type I error rate-meaning">
<summary>Common misconception: Type I error rate of 5% means a significant result has a 5% probability of being a false positive</summary>

Type I error rate of 5% means that the testing procedure rejects in about 5% of repetitions when the null hypothesis is true, under its assumptions. It does not give the probability that a significant result is a false positive. Type I error rate conditions on the null hypothesis being true; it does not tell us how likely that hypothesis is after observing the data.

</details>

Type I error rate of 5% and power of 80% or 90% are common planning choices. Choose and justify them in context. The ratio of Type II error rate to Type I error rate is not, by itself, a calculation of the relative clinical costs of the two errors. Non-rejection does not establish that the treatments are equivalent.

## See Type I error rate and Type II error rate as areas {#alpha-beta}

We now use a single-sample illustration to make the two error probabilities visible. Suppose the null mean systolic blood pressure is 120 mmHg, the specified alternative is 125 mmHg, the population SD is known to be 15 mmHg, and 25 independent observations are taken. For this **one-sided known-SD illustration**, rejection occurs when the sample mean exceeds a critical value. The two-arm treatment-comparison activities use two-sided tests.

Both curves represent the **sampling distribution of the sample mean**, not the distribution of individual blood pressures. Their SD is the standard error, $15/\sqrt{25}=3$ mmHg. Type I error rate is the rejection area under the null curve. Under the specified alternative, Type II error rate is the non-rejection area, and power is the rejection area.

```{r alpha-beta-areas}
mu0 <- 120
mu1 <- 125
sigma <- 15
n <- 25
alpha <- .05
standard_error <- sigma / sqrt(n)
critical <- mu0 + qnorm(1 - alpha) * standard_error
beta <- pnorm(critical, mean = mu1, sd = standard_error)
c(standard_error = standard_error, critical = critical,
  beta = beta, power = 1 - beta)
```

```{r alpha-beta-areas-formatted, echo=FALSE}
knitr::kable(data.frame(Measure=c("Standard error", "Critical mean", "Type I error rate", "Type II error rate", "Power"), Result=c(paste(standard_error,"mmHg"), paste(round(critical,1),"mmHg"), percent(alpha), percent(beta), percent(1-beta))),
  caption = "Error rates and power for the specified sampling distributions.")
```

```{r sampling-distribution-figure, echo=FALSE, fig.height=6, fig.cap="Sampling distributions of the mean with 25 independent observations and known SD 15 mmHg. The shaded rejection area under the null is Type I error rate; the shaded non-rejection area under the specified alternative is Type II error rate.", fig.alt="The null curve is centred at 120 and the alternative curve at 125 mmHg. Both have standard error 3. The critical mean is about 124.9; Type I error rate is 5 percent and Type II error rate about 49 percent."}
sampling_example <- normal_sampling_operating_characteristics(
  mu0 = mu0, mu1 = mu1, sigma = sigma, n = n, alpha = alpha,
  sidedness = "greater")
plot_sample_mean_distributions(sampling_example)
```

Here the critical mean is about 124.9 mmHg and power is about 51%. Repeating the study at the specified alternative would produce a mean above that threshold about half the time. This does not mean that the alternative has a 51% probability of being true.

[Explore the two sampling distributions](http://127.0.0.1:8769/study/?activity=alpha_beta). Increase n while holding the two population means and SD fixed. The curves become narrower because the standard error decreases. Explain how the rejection rule still controls Type I error rate while Type II error rate changes. Then compare one-sided and two-sided rules; sidedness should follow the research question before data collection.

## Activity 1: one study, then many

Use the chronic-pain planning probabilities: 30% relief with control and 60% with treatment, with Type I error rate of 5%, and 20, 50 and 80 patients **per group**. The simulations evaluate the 30-percentage-point planning effect; the expected benefit is 40 points and the clinical threshold is 20 points. Before looking at the results, predict which design will reject the null most often. This repeated-study approach follows the introductory teaching idea in @superpowerbook2022 and the broader use of simulation in health-sciences education [@thiesmeier2024; @rudolph2021].

```{r three-samples}
sample_sizes <- c(20, 50, 80)
batches <- lapply(sample_sizes, function(n) {
  scenario <- study_spec(outcome = "proportions", n = n,
    p_control = .30, p_treatment = .60, threshold = .20,
    width_target = .20, B = 5000, seed = 20260921 + n)
  simulate_trials(scenario)
})
power_results <- do.call(rbind, lapply(batches, power_row))
knitr::kable(display_probabilities(power_results,
  c("simulated_power", "MC_lower_95", "MC_upper_95")),
  caption = "Estimated power at a 30-percentage-point true benefit for three sample sizes. The 95% intervals describe Monte Carlo uncertainty.")
```

The following histograms count the p-values from the independent simulated studies. The dashed line marks Type I error rate. The fraction below the line estimates power under the specified alternative (Figure \@ref(fig:p-values)).

```{r p-values, echo=FALSE, fig.height=7, fig.cap="P-values from 5,000 independent clinical trials at each sample size. The proportion below Type I error rate estimates power for the assumed relief probabilities.", fig.alt="Three histograms compare p-values with 20, 50 and 80 patients per group. The numerical rejection fractions and their Monte Carlo intervals are in the preceding table."}
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (batch in batches) plot_p_values(batch)
```

::: {.activity}
**Explore.** [Open Activity 1 with its case loaded](http://127.0.0.1:8769/study/?activity=pain_one_many). In the lab choose *Binary event outcome* and *Assess a fixed sample size*. The event here is pain relief. Set 20 per group, control 30%, treatment 60%, Type I error rate 5%, and 5,000 replications. Proportion controls use 0.30, 0.60 and 0.05 for these percentages. Keep the expected benefit at 40 percentage points and clinical threshold at 20 points. Run one study, then many studies. Repeat at 50 and 80 per group.

**Explain.** Why can two studies with the same design reach different testing decisions? Why does changing the number of replications improve numerical precision without increasing the power of each study?

**Check under no effect.** Keep the plan fixed and choose *No true treatment effect*. The rejection fraction now estimates Type I error. Compare it with Type I error rate; finite-sample behaviour of a binary test can differ from its nominal level.
:::

## Precision of a simulation

Use $n$ for patients per group and $B$ for the number of simulated studies. If $\hat p$ is the simulated rejection fraction, its approximate Monte Carlo standard error is:

\begin{equation}
\mathrm{MCSE}=\sqrt{\hat p(1-\hat p)/B}.
(\#eq:mcse)
\end{equation}

At a rejection rate of 80% and $B=5000$, MCSE is about `r round(100 * sqrt(.8 * .2 / 5000), 2)` percentage points. The tables use Wilson intervals for Monte Carlo uncertainty. These intervals remain informative even when a simulation observes no rejections or all rejections. Increasing $B$ does not resolve uncertainty about the real treatment effect.

# Calculate a sample size {#calculations}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Calculate approximate sample sizes for two independent proportions or means.
- Explain how the planning effect, variability, Type I error rate and target power affect the calculation.
- Distinguish per-group from total counts.
:::

We return to two-arm treatment comparisons. The formulas below assume two independent groups with equal allocation and a two-sided test. Later, [Recruitment, precision and feasibility](#feasibility) extends planning to estimation and single-proportion examples, and [Where does my study fit?](#study-designs) considers other designs.

Having specified the chronic-pain planning effect and acceptable error risks, we can calculate the required number of **analysable patients**. We start with that trial, then consider a continuous-outcome comparison.

## The multiplier

For familiar normal approximations, define:

\begin{equation}
f(\alpha,\beta)=\left(z_{1-\alpha/2}+z_{1-\beta}\right)^2.
(\#eq:multiplier)
\end{equation}

Here $z_q$ is a standard-normal quantile. These formulas make the relationships visible; the required method ultimately depends on the planned analysis [@kirkwood2003; @wang2020].

In the equations, $\alpha$ denotes the Type I error rate and $\beta$ the Type II error rate; target power is $1-\beta$. The multiplier gathers the two error targets into one quantity. A smaller Type I error rate makes a false rejection less frequent under the null, and a higher target power requires more information at the planning effect. Both increase the multiplier and therefore the sample size when the other assumptions stay the same. You can read the table below without calculating z-scores yourself.

```{r multiplier}
alpha <- .05
target_power <- .90
f_ab <- (qnorm(1 - alpha / 2) + qnorm(target_power))^2
f_ab
```

For a Type I error rate of 5% and target power of 90%, the multiplier is **`r round(f_ab, 2)`**.

```{r multiplier-table, echo=FALSE}
grid <- expand.grid(alpha = c(.05, .01), power = c(.80, .90, .95))
grid$multiplier <- with(grid, (qnorm(1 - alpha / 2) + qnorm(power))^2)
knitr::kable(display_probabilities(grid, c("alpha", "power")), digits = 2,
  caption = "Normal-approximation sample-size multipliers for different Type I error rate and target-power choices.")
```

## Two proportions

For probabilities $\pi_T$ and $\pi_C$ in two independent, equally sized groups:

\begin{equation}
n\approx f(\alpha,\beta)
\frac{\pi_T(1-\pi_T)+\pi_C(1-\pi_C)}{(\pi_T-\pi_C)^2}.
(\#eq:proportions)
\end{equation}

The result is **per group**. Round up per group, then double for the total.

```{r proportions}
pi_t <- .60
pi_c <- .30
n_raw <- f_ab * (pi_t * (1 - pi_t) + pi_c * (1 - pi_c)) /
  (pi_t - pi_c)^2
n_per_group <- ceiling(n_raw)
c(per_group = n_per_group, total = 2 * n_per_group)
```

```{r proportions-formatted, echo=FALSE}
knitr::kable(data.frame(`Analysable per group`=n_per_group, `Analysable total`=2*n_per_group, check.names=FALSE),
  caption = "Approximate analysable sample size for the chronic-pain trial.")
```

This approximation gives `r n_per_group` per group and `r 2 * n_per_group` in total. It is not a guarantee of 90% finite-sample power. The lab uses the normal approximation implemented by `power.prop.test()` for planning and checks the chosen score test using simulation [@rcore].

<details class="misconception" id="misconception-definitive-sample">
<summary>Common misconception: the calculator provides one definitive minimum number</summary>

The calculated sample size depends on the effect, event rates or variability, error targets and analysis method you specify. It is a conditional planning result. Check plausible alternative assumptions and whether the reported number is per group or for the whole study before interpreting it.

</details>

```{r proportion-changes}
proportion_cases <- data.frame(
  scenario = c("Original", "Smaller difference", "Lower probabilities",
               "Type I error rate 1%", "Power 95%"),
  treatment = c(.60, .60, .30, .60, .60),
  control = c(.30, .45, .15, .30, .30),
  alpha = c(.05, .05, .05, .01, .05),
  power = c(.90, .90, .90, .90, .95))
proportion_cases$per_group <- mapply(sample_size_two_proportions,
  proportion_cases$treatment, proportion_cases$control,
  proportion_cases$alpha, proportion_cases$power)
proportion_cases$total <- 2 * proportion_cases$per_group
knitr::kable(display_probabilities(proportion_cases,
  c("treatment", "control", "alpha", "power")),
  caption = "Analysable sample sizes under alternative binary-outcome assumptions, using the simple normal approximation.")
```

Compare each row with the original. Halving the difference does not necessarily multiply binary sample size by exactly four: changing a probability also changes its variance. In the lower-probabilities row, treatment relief remains twice the control probability.

Read a row as a conditional statement: **if** those event probabilities and that analysis are reasonable, **then** this is the approximate number required for the specified error targets. The row is not evidence that the intervention will achieve that benefit. The smaller-difference row needs more participants because sampling variation makes that difference harder to distinguish from zero.

### An adherence-support case {#adherence}

Suppose you compare usual care with individual adherence support, using adequate adherence at 12 weeks as a yes/no outcome. Comparable evidence suggests an expected improvement from 50% to 70%, a **20-percentage-point expected benefit**. For planning, use 50% versus 65%, a more conservative **15-percentage-point benefit**. Patients and clinicians separately judge a **10-percentage-point benefit** worthwhile. These three values answer different questions.

```{r adherence-case}
adherence_case <- teaching_case("adherence")
adherence_plan <- teaching_case_spec("adherence")
adherence_plan$n <- plan_n(adherence_plan, "testing",
  adherence_case$inputs$target_power)
c(analysable_per_group = adherence_plan$n,
  recruit_per_group = adjust_for_dropout(adherence_plan$n,
    adherence_case$inputs$dropout),
  planning_power = spec_power(adherence_plan))
```

```{r adherence-case-formatted, echo=FALSE}
knitr::kable(data.frame(`Analysable per group`=adherence_plan$n, `Recruit per group`=adjust_for_dropout(adherence_plan$n,adherence_case$inputs$dropout), `Planning power`=percent(spec_power(adherence_plan)),check.names=FALSE),
  caption = "Sample size and recruitment requirements for adherence support.")
```

[Explore the adherence case](http://127.0.0.1:8769/study/?activity=adherence). Compare the recruitment requirement with a limit of 240 participants and 15% expected loss per arm. Then keep the original planned n fixed and reduce the generating percentage under support to 60%. Explain why a smaller actual benefit changes power without changing the original recruitment plan. These methods assume patients are allocated independently; allocating entire practices would require cluster methods.

Under these assumptions, the required recruitment total is `r 2 * adjust_for_dropout(adherence_plan$n, adherence_case$inputs$dropout)`, exceeding the proposed limit of 240. This is a reason to reconsider recruitment, collaboration or the study's aim. Increasing the assumed benefit merely to obtain a smaller n would not supply the missing evidence.

## Two means

For a mean difference $\delta$ and common population standard deviation $\sigma$:

\begin{equation}
n\approx f(\alpha,\beta)\frac{2\sigma^2}{\delta^2}.
(\#eq:means)
\end{equation}

Suppose comparable evidence suggests a **4-unit expected difference** in a normally distributed improvement score. Use a more conservative **3-unit planning difference**, with SD 5 in both groups, while patients and clinicians judge a **2-unit benefit clinically important**. This is a hypothetical score, not a bounded clinical scale. Keep Type I error rate at 5% and power at 90%.

```{r means}
delta <- 3
sigma <- 5
n_per_group <- ceiling(f_ab * 2 * sigma^2 / delta^2)
c(per_group = n_per_group, total = 2 * n_per_group)
```

```{r means-formatted, echo=FALSE}
knitr::kable(data.frame(`Analysable per group`=n_per_group, `Analysable total`=2*n_per_group, check.names=FALSE),
  caption = "Approximate analysable sample size for the continuous-outcome comparison.")
```

The equal-allocation total is `r 2 * n_per_group`, rather than rounding the unrounded total to 117. The t-test calculation accounts for estimating variance:

```{r t-test-planning}
t_plan <- power.t.test(delta = 3, sd = 5, sig.level = .05,
  power = .90, type = "two.sample", alternative = "two.sided",
  strict = TRUE)
t_plan
c(per_group = ceiling(t_plan$n), total = 2 * ceiling(t_plan$n))
```

```{r t-test-planning-formatted, echo=FALSE}
knitr::kable(data.frame(Method=c("Normal approximation", "Planned two-sample t test"), `Analysable per group`=c(n_per_group,ceiling(t_plan$n)), `Analysable total`=2*c(n_per_group,ceiling(t_plan$n)), check.names=FALSE),
  caption = "Sample size requirements under two planning methods.")
```

These results distinguish a transparent normal approximation from a calculation for the planned t test. A small difference in n comes from the method, rather than from a different clinical question. Report the method used and round within each group before doubling; recruitment losses are considered separately.

```{r mean-changes}
mean_cases <- data.frame(
  scenario = c("Original", "Half the difference", "Twice the SD",
               "Type I error rate 1%", "Power 95%"),
  difference = c(3, 1.5, 3, 3, 3), sd = c(5, 5, 10, 5, 5),
  alpha = c(.05, .05, .05, .01, .05),
  power = c(.90, .90, .90, .90, .95))
mean_cases$per_group <- mapply(sample_size_two_means,
  mean_cases$difference, mean_cases$sd, mean_cases$alpha, mean_cases$power)
mean_cases$total <- 2 * mean_cases$per_group
knitr::kable(display_probabilities(mean_cases, c("alpha", "power")),
  caption = "Analysable sample sizes under alternative mean differences, variability and error targets, using the simple normal approximation.")
```

Before rounding, halving the difference or doubling the SD multiplies the required sample size by four. Explain this using equation \@ref(eq:means).

### A rehabilitation case {#rehabilitation}

The same calculation could compare two rehabilitation programmes, rather than two drugs. Suppose the primary outcome is improvement at six weeks on a hypothetical continuous mobility score. The expected benefit is **4 units**, the planning difference is **3 units**, and patients and clinicians judge a **2-unit benefit** clinically meaningful. Use a common SD of 5 units. For this illustration, independent patients are randomized to the two programmes and the improvement scores are approximately normal.

```{r rehabilitation-case}
rehabilitation_case <- teaching_case("rehabilitation")
rehabilitation_plan <- teaching_case_spec("rehabilitation")
rehabilitation_plan$n <- plan_n(rehabilitation_plan, "testing",
  rehabilitation_case$inputs$target_power)
c(analysable_per_group = rehabilitation_plan$n,
  total = 2 * rehabilitation_plan$n,
  recruitment_total = 2 * adjust_for_dropout(rehabilitation_plan$n,
    rehabilitation_case$inputs$dropout))
```

```{r rehabilitation-case-formatted, echo=FALSE}
knitr::kable(data.frame(`Analysable per group`=rehabilitation_plan$n, `Analysable total`=2*rehabilitation_plan$n, `Recruit total`=2*adjust_for_dropout(rehabilitation_plan$n, rehabilitation_case$inputs$dropout), check.names=FALSE), caption="Sample size and recruitment requirements for the rehabilitation comparison.")
```

[Explore the rehabilitation case](http://127.0.0.1:8769/study/?activity=rehabilitation). Ask whether 10% expected loss and a recruitment limit of 200 make this plan feasible. Compare the confidence interval with the 2-unit clinical threshold. Although an improvement score uses before/after information within each patient, this example analyses one score per patient across **two independent groups**. A single-group paired comparison, repeated-measures model, or baseline-adjusted analysis needs its own planning method.

The expected-loss adjustment gives a recruitment total of `r 2 * adjust_for_dropout(rehabilitation_plan$n, rehabilitation_case$inputs$dropout)`, below this example's cap of 200. Meeting the cap and power target still leaves questions about the credibility of the assumptions, measurement quality and the precision needed for a clinical decision.

# Read power curves {#curves}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Read power curves.
- Compare plausible effects and sample sizes.
- Assess sensitivity and recruitment limits while keeping the planned sample size fixed when appropriate.
:::

The chronic-pain calculation uses one set of assumptions. What if the true benefit is smaller, or recruitment limits the number of patients? Power curves show how the answer changes.

## Activity 2: can the feasible study answer the question?

A single calculation hides its dependence on uncertain inputs. Chapter 11 of @superpowerbook2022 motivates examining power across a range of sample sizes and effects. Here we keep the chronic-pain plan fixed and examine true control relief of 30% with treatment relief of 45%, 50% or 60%. The case still has an expected benefit of 40 percentage points, a planning benefit of 30 points and a clinical threshold of 20 points; these curves explore performance if reality differs from the expectation.

Suppose the clinic can recruit **200 patients in total**, with 10% expected loss in each arm. This corresponds to 100 recruited and an expected 90 analysable patients per group. That expected count is not guaranteed.

```{r curve-settings}
curve_spec <- study_spec(outcome = "proportions", n = 90,
  p_control = .30, p_treatment = .60, threshold = .20,
  width_target = .20)
effect_sizes <- c(.15, .20, .30)
target_power <- .90
recruitment_cap <- 200
expected_loss <- .10
analysable_cap <- floor(recruitment_cap / 2 * (1 - expected_loss))
```

```{r power-curves, fig.cap="Approximate power versus analysable patients per group. Curves represent three absolute benefits; dashed lines mark target power and the expected analysable limit from the recruitment cap.", fig.alt="Three curves show that smaller treatment benefits need more patients. A horizontal line marks 90 percent power and a vertical line marks 90 analysable patients per group."}
curve_table <- power_curve_data(curve_spec, effect_sizes, seq(5, 200, 5))
plot_power_curves(curve_table, target_power, analysable_cap)
```

```{r cap-table, echo=FALSE}
cap_results <- do.call(rbind, lapply(effect_sizes, function(effect) {
  x <- curve_spec; x$p_treatment <- x$p_control + effect
  data.frame(absolute_benefit = effect,
    analysable_per_group = analysable_cap,
    approximate_power_at_cap = spec_power(x, analysable_cap),
    per_group_for_target = plan_n(x, target_power = target_power))
}))
knitr::kable(display_probabilities(cap_results, "approximate_power_at_cap",
  points = "absolute_benefit"),
  caption = "Approximate power with 90 analysable patients per group at different true benefits, and the sample size required for 90% power at each benefit.")
```

## Hold sample size fixed and vary the effect

Figure \@ref(fig:effect-curve) answers a different question: with 90 analysable patients per group, how does power change across possible treatment benefits? The target-power crossing describes a detectable effect at that target.

```{r effect-curve, fig.cap="Approximate power across true treatment benefits with 90 analysable patients per group and 30% control relief. The horizontal reference marks 90% power.", fig.alt="Power rises with the assumed positive treatment benefit; the horizontal reference is 90 percent power."}
effect_grid <- seq(0, .40, .005)
effect_power <- vapply(effect_grid, function(effect) {
  x <- curve_spec; x$p_treatment <- x$p_control + effect
  spec_power(x)
}, numeric(1))
plot(effect_grid * 100, effect_power, type = "l", lwd = 2,
  col = "#176675", ylim = c(0, 1), yaxt = "n",
  xlab = "Absolute treatment benefit (percentage points)", ylab = "Power (%)")
axis(2, at = seq(0, 1, .2), labels = paste0(seq(0, 100, 20), "%"))
abline(h = target_power, lty = 2)
```

<details class="misconception" id="misconception-detection-boundary">
<summary>Common misconception: the target-power crossing is a detection boundary</summary>

An effect below the target-power crossing can still produce a significant result; an effect above it can still go undetected. The curve describes probabilities across repeated studies, not a boundary separating detectable from undetectable effects. The clinical threshold is a separate judgement.

</details>

::: {.activity}
**Predict.** Which effect assumption is most demanding at the same power target?

**Explore.** [Open Activity 2 with its recruitment limit loaded](http://127.0.0.1:8769/study/?activity=pain_curves). In the lab set a fixed sample of 90 per group, control 30%, and treatment 60%. Compare true treatment relief of 50% and 45% while keeping the expected benefit, planning benefit and clinical threshold distinct. Use the prepared curves above to view all three scenarios together.

**Explain.** Which assumptions make the target feasible? How do you distinguish 200 recruited patients from 180 expected analysable patients?

**Justify.** Write three sentences describing what the clinic's recruitment limit could establish. Do not increase the assumed benefit merely to make the design affordable.
:::

## Sensitivity is not observed post hoc power

These curves specify hypothetical effects before interpreting a realised study.

<details class="misconception" id="misconception-observed-power">
<summary>Common misconception: power calculated from the observed effect explains the completed study</summary>

Substituting the study's observed effect into a power calculation adds no useful evidence about that same result. Interpret the estimate, confidence interval, study design and limitations instead. At a fixed sample size, power across independently justified hypothetical effects can still help describe the design's capabilities [@althouse2021].

</details>

For example, suppose a completed study estimates a 1-unit improvement with a 95% confidence interval from -1 to 3, while a 2-unit benefit would matter clinically. The interval includes some harm, zero, a small benefit and a clinically important benefit. The study leaves those possibilities unresolved; calculating low power from the observed 1-unit estimate does not decide between them.

With an existing records dataset, explain why that sample is available, assess data quality and missingness, and report what its estimate and interval support. A sensitivity analysis at independently justified effects can still describe a fixed design's performance. It should not use the observed effect to create additional evidence for the same study. This distinction follows the sample size justification approach in @lakens2022; [the FAQ](#common-mistakes) gives a short retrospective-study example.

# Build a simulated clinical trial {#simulation}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Generate and analyse one hypothetical trial.
- Repeat the trial to assess rejection and interval coverage.
- Explain how Monte Carlo uncertainty differs from uncertainty in the planning assumptions.
:::

This chapter returns to the two-arm chronic-pain trial and its planned analysis.

The curves in Figure \@ref(fig:power-curves) describe the performance of the chronic-pain design across repeated studies. Now generate individual trials under the same assumptions to see why their estimates, intervals and decisions vary.

## Activity 3: generate, analyse, repeat, summarise

The custom-simulation approach in chapter 15 of @superpowerbook2022 starts from the planned trial and its analysis. Our binary example uses the same score test as the existing reasoning lab. This is an adaptation of the workflow, not a reproduction of the book's logistic-regression example.

[Open Activity 3 with the trial assumptions loaded](http://127.0.0.1:8769/study/?activity=pain_simulation) to connect these four steps with one study and repeated studies in the app. The prepared batch and app use the same seed, sample size and analysis.

### Step 1: generate one trial

Generate 60 independent patients per group, with relief probabilities of 30% and 60%. The simulation evaluates the 30-percentage-point planning effect; the expected effect is 40 points and the clinical threshold is 20 points. Each patient's outcome is zero or one. Setting the seed makes this example reproducible.

```{r generate-trial}
set.seed(20260921)
n <- 60
control <- rbinom(n, size = 1, prob = .30)
treatment <- rbinom(n, size = 1, prob = .60)
c(control_relief = sum(control), treatment_relief = sum(treatment))
```

```{r generate-trial-formatted, echo=FALSE}
knitr::kable(data.frame(Group=c("Standard treatment", "Novel treatment"), Patients=n, `Reporting relief`=c(sum(control),sum(treatment)), `Relief percentage`=percent(c(mean(control),mean(treatment))), check.names=FALSE),
  caption = "Observed pain relief in one simulated trial.")
```

### Step 2: apply the planned analysis

Use a two-sided pooled score test without continuity correction. The helper also calculates a Newcombe-Wilson interval for the risk difference [@newcombe1998]. These are different procedures: test rejection and interval exclusion of zero can occasionally disagree.

```{r analyse-trial}
one_result <- analyse_binary(sum(treatment), sum(control), n, alpha = .05)
one_result
```

```{r analyse-trial-formatted, echo=FALSE}
knitr::kable(data.frame(Measure=c("Observed treatment difference", "95% confidence interval", "p-value", "Testing decision"), Result=c(paste(round(100*one_result$estimate,1),"percentage points"), paste(round(100*one_result$lower,1),"to",round(100*one_result$upper,1),"percentage points"), formatC(one_result$p_value,format="f",digits=3), if(one_result$p_value < .05) "Reject the zero-difference null" else "Do not reject the zero-difference null")),
  caption = "Estimated benefit and testing decision in one simulated trial.")
```

`estimate` is treatment minus control; `lower` and `upper` are confidence limits. The p-value tests a zero difference. The helper's score-test p-value can be checked against R:

```{r independent-check}
reference <- prop.test(c(sum(treatment), sum(control)), c(n, n),
  correct = FALSE)
c(helper_p = one_result$p_value, R_p = reference$p.value)
```

```{r independent-check-formatted, echo=FALSE}
knitr::kable(data.frame(Method=c("Tutorial helper", "R score test"), `p-value`=formatC(c(one_result$p_value,reference$p.value),format="f",digits=4), check.names=FALSE),
  caption = "Agreement between two implementations of the planned score test.")
```

This comparison checks the p-value, not the interval: `prop.test()` uses a different interval calculation. When both groups have all successes or all failures, the helper explicitly returns a p-value of one rather than an undefined statistic.

### Step 3: repeat independent trials

The efficient helper generates binomial counts instead of storing every patient's record. This is equivalent to adding up independent binary outcomes in each group. It applies the same analysis to every trial, using a seed set once before the batch.

```{r repeat-trials}
trial_spec <- study_spec(outcome = "proportions", n = 60,
  p_control = .30, p_treatment = .60, alpha = .05,
  threshold = .20, width_target = .30, B = 5000, seed = 20260921)
trials <- simulate_trials(trial_spec)
head(trials$results, 4)
```

### Step 4: summarise what the design achieves

```{r trial-summary}
knitr::kable(display_simulation_summary(simulation_summary(trials)),
  caption = "Repeated-trial operating characteristics, with Monte Carlo standard errors and 95% intervals.")
```

The rejection rate estimates power at these probabilities. Coverage is the percentage of confidence intervals containing the generating difference. The width-target rate is the percentage with a **full** interval width no greater than 30 percentage points. These are three separate properties.

<details class="misconception" id="misconception-simulation-repetitions">
<summary>Common misconception: running more simulations makes the study more powerful</summary>

Increasing the number of simulated studies makes the estimated rejection rate more precise. It does not add patients to each study or increase its underlying power. Increasing the patients per group changes the study; increasing the simulation repetitions improves our numerical description of it.

</details>

```{r trial-intervals, echo=FALSE, fig.height=6, fig.cap="Risk-difference confidence intervals from the first 30 simulated trials. Filled dots indicate rejection of zero. Reference lines mark the 20-percentage-point clinical threshold and the 30-percentage-point generating effect used for planning.", fig.alt="Intervals vary from trial to trial. Separate lines mark zero, the 20-point clinical threshold and the 30-point generating effect. The preceding table reports rejection, coverage and attainment of the full interval width target across all 5,000 trials."}
plot_trial_intervals(trials$results, trial_spec)
```

## Compare simulation with a planning approximation

```{r simulation-benchmark}
c(simulated_power = mean(trials$results$reject),
  approximate_planning_power = spec_power(trial_spec))
```

```{r simulation-benchmark-formatted, echo=FALSE}
knitr::kable(data.frame(Method=c("Repeated-trial simulation", "Planning approximation"), Power=percent(c(mean(trials$results$reject),spec_power(trial_spec)))),
  caption = "Power estimated by simulation and by the planning approximation.")
```

A difference can reflect Monte Carlo error or the approximation used for binary planning. Increasing the number of simulations reduces only the Monte Carlo component. Simulation is a way to investigate the analysis under explicit assumptions, not an automatic guarantee that those assumptions fit a real study.

::: {.activity}
**Predict.** Keep 60 patients per group. What happens if true treatment relief is 45% rather than 60%?

**Explore.** In the lab retain the planning probabilities, choose *Custom effect or variability*, set generating control to 30% and treatment to 45%, then run many studies. The planned sample must stay fixed. Repeat under no true effect.

**Explain.** Report the generating probabilities, patients per group, replications, rejection rate and its Monte Carlo interval. Distinguish uncertainty in the simulation from uncertainty in the clinical assumptions.

**Justify.** Would the original plan still answer the clinical question? State what extra evidence or design change would help.
:::

# Recruitment, precision and feasibility {#feasibility}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Adjust recruitment for expected losses.
- Interpret confidence intervals against zero and a clinical threshold.
- Plan useful precision for a prevalence estimate or pilot feasibility outcome.
- Assess what an existing dataset can establish for a new research question.
:::

The chronic-pain trial needs enough analysable outcomes, but recruitment is only part of its plan. We also need to allow for losses and consider whether the resulting estimates would guide a clinical decision.

## Account for expected losses

If $n$ analysable patients are needed in each arm and the expected dropout proportion is $r$, the simple inflation is:

\begin{equation}
n_{\mathrm{recruit,arm}}=\left\lceil\frac{n}{1-r}\right\rceil.
(\#eq:dropout)
\end{equation}

```{r recruitment}
analysable_per_group <- sample_size_two_proportions(.60, .30, power = .90)
recruit_per_group <- ceiling(analysable_per_group / (1 - .05))
c(analysable_per_group = analysable_per_group,
  recruit_per_group = recruit_per_group, recruit_total = 2 * recruit_per_group)
```

```{r recruitment-formatted, echo=FALSE}
knitr::kable(data.frame(`Analysable per group`=analysable_per_group, `Recruit per group`=recruit_per_group, `Recruit total`=2*recruit_per_group, check.names=FALSE),
  caption = "Chronic-pain recruitment target allowing for 5% expected loss.")
```

The simulations here condition on analysable counts and do not model dropout.

<details class="misconception" id="misconception-dropout">
<summary>Common misconception: adding the expected dropout percentage solves missing data</summary>

To allow for 10% expected loss, divide the required analysable count by 0.90 rather than simply adding 10%. This protects the expected participant count, but does not guarantee it. Recruiting extra participants also does not remove bias caused by missing outcomes or treatment non-adherence.

</details>

## Interpret a confidence interval clinically

For the continuous improvement-score case, the expected benefit is 4 units, the planning difference is 3 units and the clinical threshold is 2 units. Compare three hypothetical 95% confidence intervals:

| Interval | Interpretation |
|:--|:--|
| 0.2 to 2.8 | Zero is excluded, but both small and clinically important benefits remain compatible. |
| -0.8 to 3.2 | Some harm, no effect and clinically important benefit remain compatible. |
| 2.2 to 4.2 | The interval lies above the specified clinical threshold. |

Table: (\#tab:clinical-intervals) Interpreting hypothetical 95% intervals against zero and a 2-unit clinical threshold.

A confidence procedure with 95% coverage contains the fixed population effect in about 95% of repetitions under its assumptions. It does not assign a 95% probability to the fixed truth after observing one interval. Precision and clinical interpretation complement the power calculation [@sandoval2025].

<details class="misconception" id="misconception-power-precision">
<summary>Common misconception: high power guarantees clinically useful precision</summary>

A study can detect a difference from zero while leaving considerable uncertainty about whether the benefit is clinically worthwhile. Power concerns the testing decision; precision concerns the interval's width. Consider both when deciding whether the study can answer the clinical question.

</details>

## Plan for precision

The lab offers *Estimate with a target interval width* alongside testing and fixed-resource goals. Specify the **full** width, in the outcome's original units. For the continuous improvement score, keep the expected benefit at 4 units, planning difference at 3 units and clinical threshold at 2 units. Use SD 5 and a target full width of 4 units:

```{r precision-plan}
precision_spec <- study_spec(outcome = "means", delta = 3, sd = 5,
  threshold = 2, width_target = 4, B = 5000, seed = 20260922)
precision_spec$n <- plan_n(precision_spec, goal = "precision")
c(per_group = precision_spec$n,
  expected_full_width = anticipated_width(precision_spec),
  power_at_planning_effect = spec_power(precision_spec))
precision_batch <- simulate_trials(precision_spec)
knitr::kable(display_simulation_summary(simulation_summary(precision_batch)),
  caption = "Precision planning and realised interval performance are different quantities.")
```

```{r precision-plan-formatted, echo=FALSE}
knitr::kable(data.frame(`Analysable per group`=precision_spec$n, `Anticipated full interval width`=round(anticipated_width(precision_spec),2), `Power at the planning effect`=percent(spec_power(precision_spec)),check.names=FALSE), caption="Sample size, anticipated precision and power for the improvement-score comparison.")
```

For normal data, the curve gives expected full t-interval width. For binary data, it uses a plug-in Newcombe-Wilson width at the assumed probabilities, not exact expected width. Neither criterion guarantees that each realised interval will meet the target; the simulation reports the fraction that does so separately [@sandoval2025; @newcombe1998].

::: {.activity}
**Compare goals.** [Open the precision activity](http://127.0.0.1:8769/study/?activity=mean_precision). In the lab choose *Continuous outcome*, expected benefit 4, planning difference 3, SD 5, clinical threshold 2, Type I error rate 5% and full width 4. The outcome here is an improvement score. Compare the sample selected for estimation with the sample selected for 80% power. Explain why the same research question might require considering both criteria.
:::

## Estimate one proportion: a prevalence survey {#prevalence}

Here the design changes from a two-arm treatment comparison to a survey estimating one population proportion. There is no treatment or control arm; sample size is chosen for a useful margin of error.

Suppose a service wants to estimate the percentage of eligible patients reporting a symptom. It wants a 95% confidence interval with an approximate margin of error of **5 percentage points**: half-width $d=0.05$, or a full width of 10 percentage points. This is a simple independent-sample prevalence example, with one total sample rather than n per treatment arm.

For an anticipated population proportion $p_0$, the familiar normal approximation gives:

\begin{equation}
n\approx\frac{z_{1-\alpha/2}^{2}p_0(1-p_0)}{d^2}.
(\#eq:single-proportion)
\end{equation}

Here $p_0$ is an assumption used before data collection, not an observed sample result. If comparable evidence suggests a prevalence of 20% ($p_0=0.20$), the approximate requirement is 246 completed observations. If there is no useful prior estimate, 50% ($p_0=0.50$) gives the largest value of $p_0(1-p_0)$ and an approximate requirement of 385.

```{r single-proportion-precision}
prevalence_known <- proportion_precision_plan(p = .20,
  full_width = .10, confidence = .95)
prevalence_unknown <- proportion_precision_plan(p = .20,
  full_width = .10, confidence = .95, unknown_p = TRUE)
knitr::kable(data.frame(
  scenario = c("Prior estimate 20%", "Unknown: use 50% for planning"),
  planning_percentage = percent(c(prevalence_known$planning_p,
    prevalence_unknown$planning_p)),
  completed_observations = c(prevalence_known$n, prevalence_unknown$n)),
  caption = "Approximate prevalence sample sizes for a 95% interval with a 5-percentage-point margin of error.")
```

The conservative 50% choice concerns **absolute precision for one proportion under this approximation**. It is not a general substitute for unknown control and treatment percentages in a power calculation. For a rare outcome, an absolute margin of five percentage points might be unhelpfully wide; event counts and the chosen interval method also matter.

[Explore the prevalence activity](http://127.0.0.1:8769/study/?activity=prevalence). Compare margins of error of 5 and 2.5 percentage points. The latter requires about four times as many completed observations under the formula. The app shows a Wilson interval for an illustrative realised sample; its width can differ from the planning approximation. No formula here guarantees that every realised interval meets the target. Representative recruitment matters as well: a larger convenience sample does not automatically remove selection bias [@whitley2002; @wang2020].

## Reuse an existing dataset for a different question {#dataset-reuse}

A bachelor project may start with an existing dataset rather than new recruitment. Suppose a completed chronic-pain trial included **200 patients** and also recorded sleep difficulty before treatment. Your new question is: **what percentage of these trial participants reported sleep difficulty at baseline?** This differs from the trial's original question about treatment effectiveness.

Baseline sleep information is available for **180** participants; **72** reported difficulty. The observed percentage is **72/180 = 40%**. The sample size is already fixed, so assess the precision the available data provide instead of calculating how many patients to recruit.

```{r reused-dataset}
sleep_result <- single_proportion_interval(events = 72, n = 180,
  confidence = .95)
sleep_result
```

```{r reused-dataset-summary, echo=FALSE}
sleep_limits <- wilson_interval(.40, 180)
knitr::kable(data.frame(
  Measure = c("Patients in the dataset", "With baseline sleep information",
    "Reporting sleep difficulty", "Observed percentage", "95% Wilson confidence interval"),
  Result = c("200", "180", "72", "40.0%",
    paste(percent(sleep_limits[1, 1]), "to", percent(sleep_limits[1, 2])))),
  caption = "A new descriptive question answered using an existing chronic-pain trial dataset.")
```

The interval describes sampling uncertainty under the model. It does not account for possible differences between the 20 patients with missing information and those who answered. The trial's eligibility criteria also restrict whom these results describe: its participants may not represent all patients with chronic pain. Random allocation supports the original treatment comparison; it does not make this baseline sample representative of a broader population.

The original trial's power calculation does not transfer to this new question. Check that the required variables and time points are available, that reuse is permitted, and that the analysis fits the new objective. For a new association question, you would also need to consider confounding and the relevant model. Avoid calculating observed post hoc power; report the estimate, interval and limitations.

::: {.activity}
**Adapt the question.** Identify the new outcome, target population, available sample and missing observations. Would an interval of this width answer your project question? Record these decisions in the [assumptions report activity](http://127.0.0.1:8769/study/?activity=assumptions_report), using a fixed-resource estimation goal.
:::

## Stress-test the study plan

Keep the sample selected under the original assumptions fixed when considering a smaller true effect or greater variability. In the lab, change *Reality may differ from the plan* rather than replacing the original planning assumptions.

```{r fixed-plan-sensitivity}
plan <- study_spec(outcome = "means", n = 45, delta = 3, sd = 5)
smaller <- plan; smaller$delta <- 1
more_variable <- plan; more_variable$sd <- 7
scenarios <- list(plan, smaller, more_variable)
knitr::kable(data.frame(
  scenario = c("Planning scenario", "Smaller effect", "Larger SD"),
  n_per_group = vapply(scenarios, function(x) x$n, numeric(1)),
  power = percent(vapply(scenarios, spec_power, numeric(1))),
  expected_full_width = vapply(scenarios, anticipated_width, numeric(1))),
  digits = 3, caption = "Sensitivity with the original sample held fixed.")
```

## Underpowered studies: what if the study is too small for its purpose? {#what-if-the-study-is-too-small-for-its-purpose}

Low power makes inconclusive results more likely under the planning alternative. Among selected statistically significant results, effects can be exaggerated. Larger samples reduce sampling variation but do not automatically remove bias.

For the chronic-pain trial, a study with low power at the 30-percentage-point planning effect would often fail to detect that benefit even if it were real. A non-significant result would not establish that the treatments are equally effective. Examine the estimated benefit and confidence interval against zero and the 20-percentage-point clinical threshold.

## Ways to increase power {#increase-power}

Before collecting data, consider whether you can obtain more information about the treatment effect. Table \@ref(tab:increase-power) separates changes to the study from changes to its error targets.

| Option | Why it can help | What to consider |
|:--|:--|:--|
| Increase the analysable sample | More independent patients reduce sampling uncertainty at the same true effect. | Extend recruitment or collaborate across sites. Allow for expected losses and check that recruitment is feasible. |
| Improve follow-up and outcome completeness | Retaining more outcomes protects the analysable sample. | Address reasons for missing outcomes; a larger count alone does not remove missing-data bias. |
| Improve measurement | More reliable measurement can reduce unexplained variability in a continuous outcome and improve precision. | For pain relief, use a clear, consistent definition and assessment time. A different outcome needs its own assumptions and calculation. |
| Use an appropriate, efficient design and analysis | Prespecified adjustment for informative baseline measures, or a suitable paired or repeated-measures design, can improve precision. | These methods require different planning calculations. Repeated observations from one patient are not extra independent patients. |
| Keep allocation balanced when appropriate | For the equal-variance, equal-cost comparisons illustrated here, equal group sizes make efficient use of a fixed total. | Unequal costs, variances or other design requirements can justify another allocation. Recalculate for that design. |
| Accept a higher Type I error rate | A less stringent rejection rule increases power at the specified alternative. | It also increases the risk of rejecting a true null. Justify the error rate prospectively; changing it to make a result significant is not sound planning. |
| Use a one-sided test when the scientific question warrants it | Directing the rejection region to a prespecified direction can increase power for effects in that direction. | It does not test both directions in the same way. Specify and justify it before seeing the data; a smaller required sample is not sufficient justification [@moye2002]. |

Table: (\#tab:increase-power) Options for increasing power and the assumptions or tradeoffs they require.

A larger **true** benefit is easier to detect, but entering a larger assumed benefit into a calculator does not make the treatment more effective. Keep the chronic-pain expected benefit (40 percentage points), planning benefit (30 points) and clinical threshold (20 points) distinct and justified. Changing the clinical threshold alone does not increase power for a test of zero difference. Likewise, running more simulations improves the precision of the estimated power; it does not increase the study's power.

::: {.activity}
**Compare plans.** [Open the chronic-pain activity](http://127.0.0.1:8769/study/?activity=pain_one_many). Keep standard-treatment relief at 30%, novel-treatment relief at 60% and the Type I error rate at 5%. Compare 50 and 80 analysable patients per group. Which plan detects the specified benefit more often? What would obtaining the larger sample require in practice? Record the recruitment implications in your study plan.
:::

If none of these options makes the study informative and feasible, reconsider an overly broad aim, collaborate, or decide not to proceed with an efficacy study. A feasibility study should be judged against feasibility aims rather than treated as a miniature confirmatory trial [@lakens2022]. Changing the aim changes what the study can establish; it does not make the original treatment comparison more powerful. Broader eligibility can affect variability and generalisability. Clustered designs, nonparametric analyses and Bayesian planning require appropriate methods beyond the calculations illustrated here [@hayes1999; @happ2019; @pezeshk2003].

## What can you do when information is limited? {#pilot-feasibility}

Suppose a team wants to test a new rehabilitation pathway but has little evidence about the treatment effect. It can recruit 40 patients over three months for a feasibility study. Rather than assuming a large improvement to obtain an affordable efficacy calculation, it asks whether recruitment, delivery and follow-up can work in practice.

First check systematic reviews, relevant previous studies, routine service data and clinical expertise. Record the population, outcome, time point and uncertainty of each source. Agree the main trial's clinically meaningful outcome with patients and clinicians. The pilot can investigate recruitment and retention, measurement procedures, acceptability and variability; its noisy treatment estimate should not become the sole basis of the definitive study [@albers2018].

For the pilot, state the resources and the decisions they must support. Here the hypothetical cap is 40 participants, and expected completion of follow-up is 80%. If 32 of 40 complete follow-up, what would that tell the team about retention?

```{r pilot-retention}
pilot_retention <- single_proportion_interval(events = 32, n = 40,
  confidence = .95)
pilot_retention
```

```{r pilot-retention-formatted, echo=FALSE}
knitr::kable(data.frame(`Completed follow-up`="32/40 (80%)", `95% Wilson confidence interval`=paste(percent(pilot_retention$lower), "to", percent(pilot_retention$upper)), check.names=FALSE), caption="Retention and its uncertainty in a 40-participant feasibility study.")
```

The observed 80% retention still has a wide interval. Compare it with a **prospectively agreed** progression criterion and consider reasons for missing outcomes. A point estimate of 80% alone cannot establish that the main trial will retain at least 80%.

<details class="misconception" id="misconception-pilot-goal">
<summary>Common misconception: every pilot needs a treatment-effect power calculation</summary>

A pilot may primarily investigate recruitment, retention or whether procedures are workable. Its numbers should support those feasibility questions and the precision needed for decisions, rather than a universal minimum per arm. A small pilot's uncertain treatment-effect estimate should not become the sole basis for planning the definitive trial [@albers2018].

</details>

[Explore the pilot feasibility case](http://127.0.0.1:8769/study/?activity=pilot_feasibility). It starts with a fixed total of 40 and anticipated retention of 80%. Compare illustrative intervals at 40 and 80 observations; make clear whether another recruitment period or collaboration could obtain the larger sample. This is descriptive estimation of retention, not a treatment-effect power calculation.

Check relevant guidance and the current funder's requirements before writing the protocol. For examples, [NIHR HTA guidance on feasibility and pilot studies](https://njl-admin.nihr.ac.uk/document/download/2023130) distinguishes feasibility questions from definitive efficacy questions; this is an archived guidance example. The [CONSORT extension for randomized pilot and feasibility trials](https://www.bmj.com/content/355/bmj.i5239) asks for a rationale for pilot numbers and criteria for progression [@nihrfeasibility; @eldridge2016]. Reporting guidance helps you document the design; it does not supply one sample-size formula for every study.

# Test your reasoning {#exercises}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Justify a planning goal in a new clinical case.
- Assess what a fixed sample can establish.
- Turn your evidence and assumptions into a complete study-planning report.
:::

## Discharge rates {#discharge}

Standard care discharges 80% of patients by a specified time point. Suppose comparable evidence suggests an **18-percentage-point expected benefit** (98% versus 80%), use a more conservative **16-percentage-point planning benefit** (96% versus 80%), and patients judge an **8-percentage-point benefit** worthwhile. Consider a 20% relative increase, an alternative planning benefit of 10 percentage points, and a higher power target. Use the simple approximation from equation \@ref(eq:proportions).

Before revealing the answers:

1. Convert each effect to a treatment probability.
2. Predict which scenario needs the most participants.
3. Calculate the per-group and total sample sizes.
4. Explain why the analysis method must be stated, especially with probabilities near one.

<details>
<summary>Show the calculation and interpretation</summary>

```{r discharge-answers}
discharge <- data.frame(
  scenario = c("20% relative increase", "10 percentage points", "Higher power"),
  control = .80, treatment = c(.80 * 1.20, .80 + .10, .80 * 1.20),
  power = c(.80, .80, .90))
discharge$per_group <- mapply(function(p, power) {
  sample_size_two_proportions(p, .80, alpha = .05, power = power)
}, discharge$treatment, discharge$power)
discharge$total <- 2 * discharge$per_group
knitr::kable(display_probabilities(discharge, c("control", "treatment", "power")),
  caption = "Analysable discharge-study sample sizes for alternative planning benefits and power targets, using the simple normal approximation.")
```

The treatment percentages are 96%, 90% and 96%. The approximate totals are 122, 394 and 164. The latter two are even because each group is rounded up separately. The expected benefit remains 18 points and the clinical threshold remains 8 points in each comparison.

<details class="misconception" id="misconception-percentage-points">
<summary>Common misconception: percentages and percentage points are interchangeable</summary>

From a baseline of 80%, a 20% relative increase gives 96% ($0.80\times1.20=0.96$), an increase of **16 percentage points**. An increase of 20 percentage points gives 100%. State the baseline and whether the change is relative or absolute before entering event rates into a calculator.

</details>

These are teaching approximations. Compare them with score-test planning and simulated performance before interpreting the numbers as a design recommendation.

[Open the discharge activity](http://127.0.0.1:8769/study/?activity=discharge). The 80% versus 96% scenario starts with 80% target power. Consider whether an outcome so close to 100% makes the simple approximation adequate, and examine simulated performance. A shorter time-to-discharge outcome or clustered ward allocation would require a different method.

</details>

## A fixed total of 70 patients

Suppose only 70 patients can be recruited. Explain what can be learned without treating this limit as evidence that the study has adequate power. Address the intended goal, expected losses, achievable precision and plausible effects. Propose either a defensible revised design or a narrower study purpose.

## Write a sample size justification

Write 300–600 words covering the clinical question, primary outcome, evidence for assumptions, clinical threshold, planned analysis, Type I error rate, target power or precision, per-group counts, recruitment inflation, and feasibility. Include a sensitivity curve or table and one simulation result with its seed, replications and Monte Carlo interval.

Complete the [student case template](../case-studies/student_case_template.md) and use the [assessment rubric](../teaching/assessment_rubric.md) for feedback. The same ten criteria assess assumptions, methods, power, precision and communication.

Use these review questions with a collaborator:

- Are the assumptions supported and distinguished from clinical judgements?
- Does the calculation match the planned analysis?
- Are per-group, total and recruitment counts unambiguous?
- Does the sensitivity analysis keep the appropriate quantities fixed?
- Does the interpretation acknowledge what an inconclusive result could mean?

For an extended activity, write a short hypothetical study abstract that distinguishes the planned design, realised results and interpretation. This follows the broader learning sequence used in DICE [@thiesmeier2024].

## Plan your own study {#own-study}

Apply the statistical, clinical, ethical and logistic questions introduced earlier to your own project. [Open the own-study workspace](http://127.0.0.1:8769/study/?activity=own_study) and complete the [case and assumption map](../case-studies/student_case_template.md). Use hypothetical or public values for this exercise; do not put patient records into the teaching app.

1. **Describe the question and design.** State the population, comparison, outcome and time point. Check [the study-design guide](#study-designs): one independent observation per person, paired observations, clustered recruitment and prediction models have different planning needs.
2. **Choose what the study should establish.** Specify an effect to test, a useful interval width, feasibility criteria, or what an already fixed sample could establish. Record evidence sources and plausible alternatives, rather than entering unsupported numbers.
3. **Connect clinical and ethical value to the design.** Who judges the benefit worthwhile? What burden, harm or opportunity cost would participation bring? What would you do if the likely interval still included both trivial and important effects? A sample meeting a conventional power target does not answer all these questions.
4. **Assess resources and communicate a recommendation.** State recruitment rate, period, per-person cost and expected losses. Distinguish available recruitment from expected analysable counts. Explore sensitivity at the original planned n and justify proceeding, collaborating, redesigning, narrowing the aim, or not proceeding. Save the assumptions, method and explanation with the app results.

Compare your study with [the rehabilitation case](#rehabilitation), [adherence-support case](#adherence), [discharge case](#discharge) and [prevalence survey](#prevalence). A different medical setting may use the same statistical method; a different dependency structure or inferential goal may require another method even if the outcome looks similar.

```{r study-design-help, echo=FALSE, results='asis'}
cat(readLines("../student-materials/study-design-guide.md", warn = FALSE),
  sep = "\n")
```

The diagnostic, prediction and preclinical resources above are also listed in the tutorial references [@stard2015; @riley2020; @arrive2020].

```{r prerequisite-help, echo=FALSE, results='asis'}
cat(readLines("../student-materials/glossary.md", warn = FALSE), sep = "\n")
```

```{r misconceptions-help, echo=FALSE, results='asis'}
cat(readLines("../student-materials/common-mistakes.md", warn = FALSE),
  sep = "\n")
```

# Reproducibility and further reading {#appendix}

::: {.learning-goals}
**Learning outcomes.** After this section, you should be able to:

- Reproduce a calculation or simulation using its inputs, method and random seed.
- Explain why results from different software must be compared using matching assumptions and methods.
:::

The current open edition uses reproducible R plots, tables and browser apps. Compare hypotheses, tests, allocation, continuity corrections and rounding before comparing results across software. Numerical agreement is informative only when the methods and inputs match; methodological validation records belong with the source documentation.

In the chronic-pain example, $P_1=0.60$ and $P_2=0.30$. Thus $P_2-P_1=-0.30$, with absolute difference 0.30. Swapping group labels changes the sign, not the power of the same two-sided test.

## Reproduce or adapt the examples

The code uses base R and the shared clinical simulation functions. The two-proportion analysis is a pooled score test without continuity correction, with Newcombe-Wilson confidence intervals. Normal-outcome examples assume independent normal observations with a common variance and use a pooled t test. Binary analytical power is a normal approximation; simulation estimates the finite-sample rejection rate of the specified test.

The shared functions are supplied in `../R/sample_size_functions.R`. Presentation helpers are in `../R/tutorial_helpers.R`. The same clinical simulation functions underpin the existing OER lab. A compact demonstration of the repeated-trial algorithm is:

```{r algorithm, eval=FALSE}
set.seed(20260921)  # Once before the batch, not inside the repetitions
B <- 5000
n <- 60
treatment_count <- rbinom(B, n, .60)
control_count <- rbinom(B, n, .30)
results <- analyse_binary(treatment_count, control_count, n, alpha = .05)
mean(results$p_value < .05)
```

The general simulation principle is also described by @lakenscaldwell2021. That paper's factorial examples are not prerequisites for these clinical activities.

<details>
<summary>R version and package information for this build</summary>

```{r session, echo=FALSE}
sessionInfo()
```

</details>

# References {-}

## AI-use disclosure {-}

OpenAI Codex assisted with developing the website and interactive applications, reviewing code, and managing the GitHub repository. The author remains fully responsible for the content and accuracy of the teaching materials and software.

Software reference: [@openaicodex].

## Sources {-}

Sources are linked at their point of use. The Superpower book informs the presentation and activity design; the clinical examples use the methods specified in this tutorial.

::: {#refs}
:::
